A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-02-14 analysis/paper/paper.Rmd
New method for shoreline dating

The close association between Scandinavian Stone Age settlements and prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century. Shoreline dating has been fundamental to Stone Age research in Norway ever since [see for example @berg-hansen2009; @bergsvik2020; @bjerck2008; @damlien2021]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

Relative sea-level (RSL) can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015]. Variation in this relative distance follow from a range of effects [e.g. @milne2009; @mörner1976]. Of central importance here is eustasy and glacial istostasy. The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015]. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins. Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].

@@

The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shorelines, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g. @conolly2020; @conolly2006,94--97]. This was done for each year along the entirety of the curves using the standard squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase. In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

The result of running the simulations across all sites are given in Figure \@ref(fig:results). Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, most of the sites have been situated close to the shoreline when they were in use. When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. It is interesting to note the distance for some of the earliest sites, which appears somewhat high. This, however, can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature and the suggested impact of the Neolithisation process.

>>>>

The close association between Scandinavian Stone Age settlements and shifting prehistoric shorelines was becoming apparent by the end of the 19th century [@degeer1896], and was firmly established and first applied as a dating method in Sweden [@hollender1901] and Norway [@brogger1905] at the turn of the century. Shoreline dating has been fundamental to Stone Age research in Norway ever since [e.g. @mikkelsen1975; @berg-hansen2009; @bergsvik2020; @bjerck2008; @damlien2021]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

Relative sea-level (RSL) can be defined as the mean elevation of the surface of the sea relative to land, or, more formally, the difference in elevation between the geoid and the surface of the Earth as measured from the Earth's centre [@shennan2015]. Variation in this relative distance follow from a range of effects [e.g. @milne2009; @mörner1976]. Of central importance here is eustasy and istostasy. The eustatic sea-level is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself [@shennan2015]. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can bind or release large amounts of water into the oceans, as well as by changes in the Earth's crust that impact the volume of ocean basins. Istostasy, on the other hand, pertains to adjustments in the crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy, which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust, which causes its weight to be redistributed. These effects thus causes the lithosphere to either subside due to increased weight, or to rebound and lift upwards due to lower weight [@milne2015].

@@

The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shorelines, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was interpolated to each site location from the displacement curves using inverse distance weighting based on the distance between the site and the isobases of the displacement curves [e.g. @conolly2020; @conolly2006,94--97]. This was done for each year along the entirety of the curves weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if they overlapped with 99.7 % probability, meaning these were effectively taken to represent the same settlement phase. In the case where there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal and then summed. Multiple phases at a single site were treated as independent of each other.

@@

Overall, as is indicated by the measures for central tendency and the almost solid line along the 0 mark on the y-axes, the simulations show that most of the sites have been situated close to the shoreline when they were in use (Figure \@ref(fig:results)). When including all dates, also those that were believed to be erroneous in the original reports, some of the sites are situated considerable distances from the shoreline. However, if one accepts the interpretation in the reports that these dates do not date the main occupation of the sites, as indicated by the artefact inventories, the second row of Figure \@ref(fig:results) gives considerable support to the notion that the sites were in use when they were situated on or close to the contemporaneous shoreline. The distances for some of the earliest sites appears somewhat high, but this can likely be explained as the result of the steepness of the displacement curves for the earliest part of the Holocene (Figure \@ref(fig:map-iso)B), which leads the uncertainty in the ^14^C-dates to give a wider possible elevation range for the sea-level. Another immediately striking result is the apparent deviation from the shoreline towards the end of the Stone Age. From around 2500 BCE several sites are situated a considerable distance from the shoreline, and while a couple remain horizontally and topographically close, most appear to be elevated a considerable distance from the sea-level, as indicated on the plot for vertical distance. This corresponds well with the literature and the suggested impact of the Neolithisation process.
